An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-29 article/article.tex
finished all paper

%% - \Abstract{} almost as usual
\Abstract{Performing inference in statistical models with an
  intractable likelihood is challenging, therefore, most
  likelihood-free inference (LFI) methods encounter accuracy and
  efficiency limitations. In this paper, we present the implementation
  of the LFI method Robust Optimisation Monte Carlo (ROMC) in the
  \proglang{Python} package \pkg{ELFI}. ROMC is a novel and efficient
  (highly-parallelisable) LFI framework that provides accurate
  weighted samples from the posterior. Our implementation can be used
  in two ways. First, a scientist may use it as an out-of-the-box LFI
  algorithm; we provide an easy-to-use API harmonised with the
  principles of \pkg{ELFI}, enabling effortless comparisons with the
  rest of the methods included in the package. Additionally, we have
  carefully split ROMC into isolated components for supporting
  extensibility. A researcher may experiment with novel method(s) for
  solving part(s) of ROMC without reimplementing everything from
  scratch. In both scenarios, the ROMC parts can run in a
  fully-parallelised manner, exploiting all CPU cores. We also provide
  helpful functionalities for (i) inspecting the inference process and
  (ii) evaluating the obtained samples. Finally, we test the
  robustness of our implementation on some typical LFI examples.}

@@

Optimization Monte Carlo (OMC), proposed by~\citet{Meeds2015}, is a
novel LFI approach. The central idea is to convert the stochastic
data-generating mechanism into a set of deterministic optimization
processes. Afterwards, \citet{Forneron2016} described a similar method
under the name `reverse sampler'. In their work,~\citet{Ikonomov2019},
identified some critical limitations of OMC and proposed Robust OMC
(ROMC) as an improved version of OMC with appropriate
modifications.

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. The implementation has been designed to
ensure extensibility. ROMC is an LFI framework; it defines a sequence
of algorithmic steps for approximating the posterior without enforcing
a specific algorithm for solving each step. Therefore, a researcher
may use ROMC as the backbone algorithm and develop novel methods to
solve each separate step.\footnote{For being a ready-to-use
  algorithm,~\citet{Ikonomov2019} proposed a default method for each
  step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
evaluation in three steps. First, for securing that our implementation
is accurate, we test it against an artificial example with a tractable
likelihood. The artificial example also serves as a step-by-step guide
for showcasing how to use the various functionalities of our
implementation. Second, we use the second-order moving average (MA2)
example from the \pkg{ELFI} package, using as ground truth the samples
obtained with Rejection ABC~\citet{lintusaari2017} using a very high
number of samples. Finally, we present the execution times of ROMC,
measuring the speed-up achieved by the parallel version of the
implementation.

@@

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism, where we can sample data points but we
cannot evaluate the likelihood. Formally, a simulator-based model is a
parameterized family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these cases, we can only access the
simulator \( m_r(\thetab) \), i.e., the black-box mechanism (computer
code) that generates samples \(\yb\) in a stochastic manner from a set
of parameters \(\thetab\). We denote the process of obtaining samples
from the simulator with \( m_r(\thetab) \rightarrow \yb \). As shown
by~\citet{Meeds2015}, it is feasible to isolate the randomness of the
simulator by introducing a set of nuisance random variables denoted by
\(\ub \sim p(\ub)\). Therefore, for a specific tuple
\((\thetab, \ub)\) the simulator becomes a deterministic mapping
\(g\), such that \(\yb=\simulator(\thetab,\ub)\). In terms of computer
code, the randomness of a random process is governed by the global
seed. Although each package may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, setting the initial seed to
a specific integer converts the simulation to a deterministic piece of
code.

As stated at the Introduction, the modeling freedom of simulator-based
models comes at the price of difficulties in inferring the parameters
of interest. Denoting the observed data as \(\data\), the main
difficulty lies at the intractability of the likelihood function
\(L(\thetab) = p(\data|\thetab)\). To better see the sources of the
intractability, and to address them, we go back to the basic
characterization of the likelihood as the (rescaled) probability of
generating data \(\yb\) that is similar to the observed data
\(\data\), using parameters \(\thetab\). Formally, the likelihood
\(L(\thetab)\) is:

\begin{equation}
  \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around
\(\data\) that is defined through a distance function \(d\), i.e., \
\(\region(\data) := \{\yb: d(\yb, \data) \leq \epsilon
\}\). Equation~\ref{eq:likelihood} highlights the main source of
intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data) | \thetab) \) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit where
\(\epsilon \to 0\). Hence, the constraint is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

\begin{equation}
  \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

@@

The approximation in
Equations~\ref{eq:approx_likelihood},~\ref{eq:approx_posterior} is by
no means the only strategy to deal with the intractability of the
likelihood function in Equation~\ref{eq:likelihood}. Other strategies include
modeling the (stochastic) relationship between \(\thetab\) and
\(\yb\), and its reverse, or framing likelihood-free inference as a
ratio estimation problem, see for example \citet{blum2010, Wood2006,
  Papamakarios2016, Papamakarios2019, Chen2019, Thomas2020,
  Hermans2020}. However, both OMC and robust OMC, which we introduce
next, are based on the approximation in Equation~\ref{eq:approx_likelihood}.

@@

As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterizing the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realization \(\ub_i \sim p(\ub)\), i.e.,\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citep{Meeds2015,
  Ikonomov2019}.

@@

Each acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). The acceptance regions \(\accregioni\) can
contain any number of disjoint subsets in the \(D\)-dimensional space
and any of these subsets can take any arbitrary shape. We should make
three important remarks. First, since the bounding boxes are built
around \(\thetab_i^*\), we focus only on the connected subset of
\(\accregioni\) that contains \(\thetab_i^*\), which we denote as
\(\accregionitheta\). Second, we want to ensure that the bounding box
\(\accregionihat\) is big enough to contain on its interior all the
volume of \(\accregionitheta\). Third, we want \(\accregionihat\) to
be as tight as possible to \(\accregionitheta\) to ensure high
acceptance rate on the importance sampling step that
follows. Therefore, the bounding boxes are built in two steps. First,
we compute their axes \(\mathbf{v}_m\), for \(m = \{1, \ldots, D\}\)
based on the (estimated) curvature of the distance at \(\thetab_i^*\),
and, second, we apply a line-search method along each axis to
determine the size of the box. We refer the reader to
Algorithm~\ref{alg:region_construction} for the details. After the
bounding boxes construction, a uniform distribution \(q_i\) is defined
on each bounding box, and is used as the proposal region for
importance sampling.

@@

For computing the weight \(w_{ij}\) in Equation~\ref{eq:sampling}, we
need to check whether the samples \(\thetab_{ij}\) drawn from the
bounding boxes are inside the acceptance region \(\accregioni\). This
can be considered to be a safety-mechanism that corrects for any
inaccuracies in the construction of \(\accregionihat\) above. However,
this check involves evaluating the distance function
\(d_i(\thetab_{ij})\), which can be expensive if the model is
complex. \citet{Ikonomov2019} proposed fitting a surrogate model
\(\tilde{d}_i(\thetab)\) of the distance function \(d_i(\thetab)\), on
data points that lie inside \(\accregionihat\). In principle, any
regression model can be used as surrogate model. \citet{Ikonomov2019}
used a simple quadratic model because it has ellipsoidal isocontours,
which facilitates replacing the bounding box approximation of
\(\accregioni\) with a tighter-fitting ellipsoidal approximation.

The training data for the quadratic model is obtained by sampling
\(\thetab_{ij} \sim q_i\) and accessing the distances
\(d_i(\thetab_{ij})\). The generation of the training data adds an
extra computational cost, but leads to a significant speed-up when
evaluating the weights \(w_{ij}\). Moreover, the extra cost is largely
eliminated if Bayesian Optimization with a Gaussian process (GP)
surrogate model \(\hat{d}_i(\thetab)\) was used to obtain
\(\thetab_i^*\) in the first step. In this case, we can use
\(\hat{d}_i(\thetab)\) instead of \(d_i(\thetab)\) to generate the
training data. This essentially replaces the global GP model with a
simpler local quadratic model which is typically more robust.

@@

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions \(\hat{\accregioni}\), which expands to three
mandatory tasks; (a) sample the nuisance variables
\(\ub_i \sim p(\ub)\) for defining the deterministic distance
functions \(d_i(\thetab)\) (Steps 3-5), (b) solve the optimization
problems for obtaining \(\thetab_i^*, d_i^*\) and keep the solutions
inside the threshold \(\epsilon\)(Steps 6-9), and (c) estimate the
bounding boxes \(\accregionihat\) to define uniform distributions
\(q_i\) on them (Step 10). Optionally, a surrogate model
\(\tilde{d}_i\) can be fitted for a faster inference phase.

If \(d_i(\thetab)\) is differentiable, using a gradient-based method
is advised for obtaining \(\thetab_i^*\) faster. In this case, the
gradients \(\nabla_{\thetab} d_i\) gradients are approximated
automatically with finite-differences. This approximation requires two
evaluations of \(d_i\) for each parameter
\(\theta_m, m \in \{1, \ldots, D\}\), which scales well only in
low-dimensional problems. If \(d_i(\thetab)\) is not differentiable,
Bayesian Optimization can be used as an alternative. In this scenario,
the training part becomes slower due to fitting of the surrogate model
and the blind optimization steps.

After obtaining the optimal points \(\thetab^*_i\), we estimate the
proposal regions. Algorithm~\ref{alg:region_construction} describes
the line search approach for finding the region boundaries. The axes
of each bounding box \(\vb_m, m = \{1, \ldots, D\}\) are defined as
the directions with the highest curvature at \(\thetab_i^*\) computed
by the eigenvalues of the Hessian matrix \(\hessian_i\) of \(d_i\) at
\(\thetab_i\) (Step 1). Depending on the algorithm used in the
optimization step, we either use the real distance \(d_i\) or the
Gaussian Process approximation \(\hat{d}_i\). When the distance
function is the Euclidean distance (default choice), the Hessian
matrix can be also computed as \(\hessian_i = \jac_i^T\jac_i\), where
\(\jac_i\) is the Jacobian matrix of the summary statistics
\(\Phi(g(\thetab, \ub_i))\) at \(\thetab_i^*\). This approximation has
the computational advantage of using only first-order
derivatives. After defining the axes, we search for the bounding box
limits with a line step algorithm (Steps 2-21). The key idea is to
take long steps \(\eta\_\text{start}\) until crossing the boundary
where \(d > \epsilon\) and then take small steps backwards to find the
exact boundary position.

@@

The inference part includes one or more of the following three tasks;
(a) sample from the posterior distribution
\( \thetab_i \sim p_{d, \epsilon}(\thetab|\data)\)
(Equation~\ref{eq:sampling}), (b) compute an expectation
\(\E_{\thetab|\data}[h(\thetab)]\) (Equation~\ref{eq:expectation})
and/or (c) evaluate the unnormalized posterior
\(p_{d, \epsilon}(\thetab|\data)\)
(Equation~\ref{eq:approx_posterior_omc}). Sampling is performed by
getting \(n_2\) samples from each proposal distribution \(q_i\). For
each sample \(\thetab_{ij}\), the distance function\footnote{As
  before, if a surrogate model \(\hat{d}\) is available, it can be
  utilized as the distance function.} is evaluated for checking if it
lies inside the acceptance region. Algorithm~\ref{alg:sampling_GB}
defines the steps for computing a weighted sample. After we obtain
weighted samples, computing the expectation is straightforward using
Equation~\ref{eq:expectation}. Finally, evaluating the unnormalized
posterior at a specific point \(\thetab\) requires access to the
distance functions \(d_i\) and the prior distribution
\(p(\thetab)\). Following Equation~\ref{eq:approx_posterior_omc}, the
evaluation involves counting how many distance functions rating over
all \(d_i(\thetab) < \epsilon\). It is worth noticing that for
evaluating the unnormalized posterior, there is no need for solving
the optimization problems and building the proposal regions.

\begin{algorithm}[H]
    \centering
    \caption{Sampling. Requires a function of distance \(d_i\), the prior distribution \(p(\thetab)\), the proposal distribution \(q_i\)}\label{alg:sampling_GB}
    \begin{algorithmic}[1]
      \State \(\thetab_{ij} \sim q_i\)
          \If {\(d_i(\thetab_{ij}) > \epsilon\)}
            \State Go to 2 \Comment{Reject sample}
          \Else {}
            \State \(w_{ij} = \frac{p(\thetab_{ij})}{q(\thetab_{ij})}\) \Comment{Compute weight}
            \State Store \((w_{ij}, \thetab_{ij})\) \Comment{Store weighted sample}
          \EndIf
    \end{algorithmic}
\end{algorithm}

@@

The overview of our implementation is illustrated in
Figure~\ref{fig:romc_overview}. Following \proglang{Python}'s naming
principles, the methods starting with an underscore (green rectangles)
represent internal (private) functions, whereas the rest (blue
rectangles) are the methods exposed at the API. In
Figure~\ref{fig:romc_overview}, it can be easily observed that the
implementation follows Algorithm~\ref{alg:romc_algorithm}. The
training part includes all the steps until the computation of the
proposal regions, i.e., sampling the nuisance variables, defining the
optimization problems, solving them, constructing the regions and
fitting local surrogate models. The inference part comprises of
evaluating the unnormalized posterior (and the normalized when is
possible), sampling and computing an expectation. We also provide some
utilities for inspecting the training process, such as plotting the
histogram of the final distances or visualizing the constructed
bounding boxes. Finally, in the evaluation part, we provide two
methods for evaluating the inference; (a) computing the Effective
Sample Size (ESS) of the samples and (b) measuring the divergence
between the approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method,
  e.g., Rejection Sampling, and we want to measure the discrepancy
  between the two approximations.}

\subsubsection*{Parallelizing ROMC method}

As discussed, ROMC has the significant advantage of being fully
parallelisable. We exploit this fact by implementing a parallel
version of the major fitting components; (a) solving the optimization
problems, (b) constructing bounding box regions. We parallelize these
processes using the built-in \proglang{Python} package
\pkg{multiprocessing}. The specific package enables concurrency, using
sub-processes instead of threads, for side-stepping the Global
Interpreter (GIL). For activating the parallel version of the
algorithm, the user simply has to set \code{elfi.ROMC(<output_node>,
  parallelize = True)}.

>>>>

%% - \Abstract{} almost as usual
\Abstract{Performing inference in statistical models with an
  intractable likelihood is challenging, therefore, most
  likelihood-free inference (LFI) methods encounter accuracy and
  efficiency limitations. In this paper, we present the implementation
  of the LFI method Robust Optimisation Monte Carlo (ROMC) in the
  \proglang{Python} package \pkg{ELFI}. ROMC is a novel and efficient
  (highly-parallelizable) LFI framework that provides accurate
  weighted samples from the posterior. Our implementation can be used
  in two ways. First, a scientist may use it as an out-of-the-box LFI
  algorithm; we provide an easy-to-use API harmonized with the
  principles of \pkg{ELFI}, enabling effortless comparisons with the
  rest of the methods included in the package. Additionally, we have
  carefully split ROMC into isolated components for supporting
  extensibility. A researcher may experiment with novel method(s) for
  solving part(s) of ROMC without reimplementing everything from
  scratch. In both scenarios, the ROMC parts can run in a
  fully-parallelized manner, exploiting all CPU cores. We also provide
  helpful functionalities for (i) inspecting the inference process and
  (ii) evaluating the obtained samples. Finally, we test the
  robustness of our implementation on some typical LFI examples.}

@@

Optimization Monte Carlo (OMC), proposed by~\citet{Meeds2015}, is a
novel LFI approach. The central idea is to convert the stochastic
data-generating mechanism into a set of deterministic optimization
processes. Afterwards, \citet{Forneron2016} described a similar method
under the name `reverse sampler'. In their work,~\citet{Ikonomov2019}
identified some critical limitations of OMC, so they proposed Robust
OMC (ROMC) an improved version of OMC with appropriate modifications.

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (\pkg{Engine for likelihood-free
  inference})~\citet{1708.00707}. The implementation has been designed
to ensure extensibility. ROMC is an LFI framework; it defines a
sequence of algorithmic steps for approximating the posterior without
enforcing a specific algorithm for solving each step. Therefore, a
researcher may use ROMC as the backbone algorithm and develop novel
methods to solve each separate step. For being a ready-to-use
algorithm,~\citet{Ikonomov2019} proposed a default method for each
step, but this choice is by no means restrictive. We have designed our
software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
illustration and the evaluation of our implementation in three
steps. First, for securing that our implementation is accurate, we
test it against an artificial example with a tractable likelihood. The
artificial example also serves as a step-by-step guide for showcasing
how to use the various functionalities of our implementation. Second,
we use the second-order moving average (MA2) example from the
\pkg{ELFI} package, using as ground truth the samples obtained with
Rejection ABC~\citet{lintusaari2017} using a very high number of
samples. Finally, we present the execution times of ROMC, measuring
the speed-up achieved by the parallel version of the
implementation.

The code of the implementation is available at the official \pkg{ELFI} \href{https://github.com/elfi-dev/elfi}{repository}. Apart
from the examples presented in the paper, there are five \pkg{Google
  Colab} notebooks available online, with end-to-end examples
illustrating:
(i) \href{https://colab.research.google.com/drive/1lGRp0XrNfZ64NN0ASB_tYEKowXwlveDC?usp=sharing}{ROMC on a synthetic \(1D\) example},
(ii) \href{https://colab.research.google.com/drive/1Fof_WmCi1YizzSI_63aEsbLXsno5gSZ3?usp=sharing}{ROMC on a synthetic \(2D\) example},
(iii) \href{https://colab.research.google.com/drive/1nkdACQ370SSc0KB1bHv4sBRaxMlMqoNH?usp=sharing}{ROMC on the Moving Average example},
(iv) \href{https://colab.research.google.com/drive/1_jHVxPSH3XcNOORZJpLU0SPzs0PF8CQ5?usp=sharing}{how to extend ROMC with a Neural Network as a surrogate model},
(v) \href{https://colab.research.google.com/drive/1RzB-V1QueP1y1nyzv_VOqR1nVz3DUH3v?usp=sharing}{how to extend ROMC with a custom proposal region module}.

@@

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism, where we can sample data points but we
cannot evaluate the likelihood. Formally, a simulator-based model is a
parameterized family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these cases, we can only access the
simulator \( m_r(\thetab) \), i.e., the black-box mechanism (computer
code) that generates samples \(\yb\) in a stochastic manner from a set
of parameters \(\thetab \in \mathbb{R}^D\). We denote the process of obtaining samples
from the simulator with \( m_r(\thetab) \rightarrow \yb \). As shown
by~\citet{Meeds2015}, it is feasible to isolate the randomness of the
simulator by introducing a set of nuisance random variables denoted by
\(\ub \sim p(\ub)\). Therefore, for a specific tuple
\((\thetab, \ub)\) the simulator becomes a deterministic mapping
\(g\), such that \(\yb=\simulator(\thetab,\ub)\). In terms of computer
code, the randomness of a random process is governed by the global
seed. Although each package may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, setting the initial seed to
a specific integer converts the simulation to a deterministic piece of
code.

The modeling freedom of simulator-based models comes at the price of
difficulties in inferring the parameters of interest. Denoting the
observed data as \(\data\), the main difficulty lies at the
intractability of the likelihood function
\(L(\thetab) = p(\data|\thetab)\). To better see the sources of the
intractability, and to address them, we go back to the basic
characterization of the likelihood as the (rescaled) probability of
generating data \(\yb\) that is similar to the observed data
\(\data\), using parameters \(\thetab\). Formally, the likelihood
\(L(\thetab)\) is:

\begin{equation}
  \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \thetab)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around
\(\data\) that is defined through a distance function \(d\), i.e., \
\(\region(\data) := \{\yb: d(\yb, \data) \leq \epsilon \}\). In cases
where the output \(\yb\) belongs to a high dimensional space, it is
common to extract summary statistics \(\Phi\) before applying the
distance \(d\). In these cases, the \(\epsilon\)-region is defined as
\(\region(\data) := \{\yb: d(\Phi(\yb), \Phi(\data)) \leq \epsilon
\}\). In our notation, the summary statistics are sometimes omitted
for simplicity. Equation~\ref{eq:likelihood} highlights the main
source of intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data) | \thetab) \) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit where
\(\epsilon \to 0\). Hence, the constraint is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

\begin{equation}
  \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \thetab), \quad \text{where  } \epsilon > 0.
\end{equation}

@@

Equations~\ref{eq:approx_likelihood} and~\ref{eq:approx_posterior} is by
no means the only strategy to deal with the intractability of the
likelihood function in Equation~\ref{eq:likelihood}. Other strategies include
modeling the (stochastic) relationship between \(\thetab\) and
\(\yb\), and its reverse, or framing likelihood-free inference as a
ratio estimation problem, see for example \citet{blum2010, Wood2006,
  Papamakarios2016, Papamakarios2019, Chen2019, Thomas2020,
  Hermans2020}. However, both OMC and robust OMC, which we introduce
next, are based on the approximation in Equation~\ref{eq:approx_likelihood}.

@@

As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterizing the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realization \(\ub_i \sim p(\ub)\), i.e.,\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citet{Meeds2015,
  Ikonomov2019}.

@@

Each acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). The acceptance regions \(\accregioni\) can
contain any number of disjoint subsets in the \(D\)-dimensional space
and any of these subsets can take any arbitrary shape. We should make
three important remarks. First, since the bounding boxes are built
around \(\thetab_i^*\), we focus only on the connected subset of
\(\accregioni\) that contains \(\thetab_i^*\), which we denote as
\(\accregionitheta\). Second, we want to ensure that the bounding box
\(\accregionihat\) is big enough to contain on its interior all the
volume of \(\accregionitheta\). Third, we want \(\accregionihat\) to
be as tight as possible to \(\accregionitheta\) to ensure high
acceptance rate on the importance sampling step that
follows. Therefore, the bounding boxes are built in two steps. First,
we compute their axes \(\mathbf{v}_m\), for \(m = \{1, \ldots, D\}\)
based on the (estimated) curvature of the distance at \(\thetab_i^*\),
and, second, we apply a line-search method along each axis to
determine the size of the bounding box. We refer the reader to
Algorithm~\ref{alg:region_construction} for the details. After the
bounding boxes construction, a uniform distribution \(q_i\) is defined
on each bounding box, and is used as the proposal region for
importance sampling.

@@

For computing the weight \(w_{ij}\) at Equation~\ref{eq:sampling}, we
need to check whether the samples \(\thetab_{ij}\), drawn from the
bounding boxes, are inside the acceptance region \(\accregioni\). This
can be considered to be a safety-mechanism that corrects for any
inaccuracies in the construction of \(\accregionihat\) above. However,
this check involves evaluating the distance function
\(d_i(\thetab_{ij})\), which can be expensive if the model is
complex. \citet{Ikonomov2019} proposed fitting a surrogate model
\(\tilde{d}_i(\thetab)\) of the distance function \(d_i(\thetab)\), on
data points that lie inside \(\accregionihat\). In principle, any
regression model can be used as surrogate model. \citet{Ikonomov2019}
used a simple quadratic model because it has ellipsoidal isocontours,
which facilitates replacing the bounding box approximation of
\(\accregioni\) with a tighter-fitting ellipsoidal approximation.

The training data for the quadratic model is obtained by sampling
\(\thetab_{ij} \sim q_i\) and accessing the distances
\(d_i(\thetab_{ij})\). The generation of the training data adds an
extra computational cost, but leads to a significant speed-up when
evaluating the weights \(w_{ij}\). Moreover, the extra cost is largely
eliminated if Bayesian optimization with a Gaussian process (GP)
surrogate model \(\hat{d}_i(\thetab)\) was used to obtain
\(\thetab_i^*\) in the first step. In this case, we can use
\(\hat{d}_i(\thetab)\) instead of \(d_i(\thetab)\) to generate the
training data. This essentially replaces the global GP model with a
simpler local quadratic model which is typically more robust.

@@

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions \(\hat{\accregioni}\), which expands to three
mandatory tasks; (a) sample the nuisance variables
\(\ub_i \sim p(\ub)\) for defining the deterministic distance
functions \(d_i(\thetab)\) (Steps 3-5), (b) solve the optimization
problems for obtaining \(\thetab_i^*, d_i^*\) and keep the solutions
inside the threshold \(\epsilon\) (Steps 6-9), and (c) estimate the
bounding boxes \(\accregionihat\) to define uniform distributions
\(q_i\) on them (Step 10). Optionally, a surrogate model
\(\tilde{d}_i\) can be fitted for a faster inference phase (Step 11).

If \(d_i(\thetab)\) is differentiable, using a gradient-based method
is advised for obtaining \(\thetab_i^*\) faster. In this case, the
gradients \(\nabla_{\thetab} d_i\) gradients are approximated
automatically with finite-differences, if they are not provided in
closed-form by the user. Finite-differences approximation requires two
evaluations of \(d_i\) for each parameter
\(\theta_m, m \in \{1, \ldots, D\}\), which scales well only in
low-dimensional problems. If \(d_i(\thetab)\) is not differentiable,
Bayesian Optimization can be used as an alternative. In this scenario,
the training part becomes slower due to fitting of the surrogate model
and the blind optimization steps.

After obtaining the optimal points \(\thetab^*_i\), we estimate the
proposal regions. Algorithm~\ref{alg:region_construction} describes
the line search approach for finding the region boundaries. The axes
of each bounding box \(\vb_m, m = \{1, \ldots, D\}\) are defined as
the directions with the highest curvature at \(\thetab_i^*\) computed
by the eigenvalues of the Hessian matrix \(\hessian_i\) of \(d_i\) at
\(\thetab_i\) (Step 1). Depending on the algorithm used in the
optimization step, we either use the real distance \(d_i\) or the
Gaussian Process approximation \(\hat{d}_i\). When the distance
function is the Euclidean distance (default choice), the Hessian
matrix can be also computed as \(\hessian_i = \jac_i^T\jac_i\), where
\(\jac_i\) is the Jacobian matrix of the summary statistics
\(\Phi(g(\thetab, \ub_i))\) at \(\thetab_i^*\). This approximation has
the computational advantage of using only first-order
derivatives. After defining the axes, we search for the bounding box
limits with a line step algorithm (Steps 2-21). The key idea is to
take long steps \(\eta\_\text{start}\) until crossing the boundary and
then take small steps backwards to find the exact boundary position.

@@

The inference part includes one or more of the following three tasks;
(a) sample from the posterior distribution
\( \thetab_i \sim p_{d, \epsilon}(\thetab|\data)\)
(Equation~\ref{eq:sampling}), (b) compute an expectation
\(\E_{\thetab|\data}[h(\thetab)]\) (Equation~\ref{eq:expectation})
and/or (c) evaluate the unnormalized posterior
\(p_{d, \epsilon}(\thetab|\data)\)
(Equation~\ref{eq:approx_posterior_omc}). Sampling is performed by
getting \(n_2\) samples from each proposal distribution \(q_i\). For
each sample \(\thetab_{ij}\), the distance function\footnote{As
  before, if a surrogate model \(\hat{d}\) is available, it can be
  utilized as the distance function.} is evaluated for checking if it
lies inside the acceptance region. Algorithm~\ref{alg:sampling_GB}
defines the steps for computing a weighted sample. After we obtain
weighted samples, computing the expectation is straightforward using
Equation~\ref{eq:expectation}. Finally, evaluating the unnormalized
posterior at a specific point \(\thetab\) requires access to the
distance functions \(d_i\) and the prior distribution
\(p(\thetab)\). Following Equation~\ref{eq:approx_posterior_omc}, we
simply count for how many deterministic distance functions it holds
that \(d_i(\thetab) < \epsilon\). It is worth noticing that for
evaluating the unnormalized posterior, there is no need for solving
the optimization problems and building the proposal regions.

\begin{algorithm}[H]
    \centering
    \caption{Sampling. Requires a function of distance \(d_i\), the prior distribution \(p(\thetab)\), the proposal distribution \(q_i\)}\label{alg:sampling_GB}
    \begin{algorithmic}[1]
      \State \(\thetab_{ij} \sim q_i \forall i\) \Comment{Sample parameters}
      \For {\(i \gets 1 \textrm{ to } n_1\)}
      \For {\(j \gets 1 \textrm{ to } n_2\)}
          \If {\(d_i(\thetab_{ij}) \leq \epsilon\)} \Comment{Accept sample}
            \State \(w_{ij} = \frac{p(\thetab_{ij})}{q(\thetab_{ij})}\) \Comment{Compute weight}
            \State Store \((w_{ij}, \thetab_{ij})\) \Comment{Store weighted sample}
            \EndIf
            \EndFor
    \EndFor
    \end{algorithmic}
\end{algorithm}

@@

The overview of our implementation is illustrated in
Figure~\ref{fig:romc_overview}. Following \proglang{Python} naming
principles, the methods starting with an underscore (green rectangles)
represent internal (private) functions, whereas the rest (blue
rectangles) are the methods exposed at the API. In
Figure~\ref{fig:romc_overview}, it can be observed that the
implementation follows Algorithm~\ref{alg:romc_algorithm}. The
training part includes all the steps until the computation of the
proposal regions, i.e., sampling the nuisance variables, defining the
optimization problems, solving them, constructing the regions and
fitting local surrogate models. The inference part comprises of
evaluating the unnormalized posterior (and the normalized when is
possible), sampling and computing an expectation. We also provide some
utilities for inspecting the training process, such as plotting the
histogram of the final distances or visualizing the constructed
bounding boxes. Finally, in the evaluation part, we provide two
methods for evaluating the inference; (a) computing the Effective
Sample Size (ESS) of the samples and (b) measuring the divergence
between the approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method,
  e.g., Rejection Sampling, and we want to measure the discrepancy
  between the two approximations.}

\subsubsection*{Parallel version of ROMC}

As discussed, ROMC has the significant advantage of being fully
parallelisable. We exploit this fact by implementing a parallel
version of the major fitting components; (a) solving the optimization
problems, (b) constructing bounding box regions. We parallelize these
processes using the built-in \proglang{Python} package
\pkg{multiprocessing}. The specific package enables concurrency, using
sub-processes instead of threads, for side-stepping the Global
Interpreter (GIL). For activating the parallel version of the
algorithm, the user simply has to set \\
\code{elfi.ROMC(<output_node>, parallelize = True)}.
